---
title: "HW5"
author: "Oryn Yehoshua"
date: "2024-10-21"
output:
  pdf_document: default
  html_document: default
---

# Question 1 

## Part A 
```{r}
getwd()
dataa<- read.csv("hw05pr01 (2).csv")
X1 <- dataa$age 
X2 <- dataa$illness
X3 <- dataa$anxiety
Y <- dataa$satisfaction
modela <- lm(satisfaction ~ age + illness + anxiety, data = dataa)
summary(modela)
```

$\hat{Y}$ = -1.14(age) - 0.44(illness) - 13.47(anxiety) + 158.49

For b1, we can interpret it to mean that for every one year increase in age, patient satsifaction decreases by 1.14 on average. 

For b2, we can interpret it to mean that for every one unit increase in the severity of illness, patient satisfaction decreases by 0.44 points on average. 
For b3, we can interpret it to mean that for every one unit increase in patient anxiety, patient satisfaction decreases by 13.47 on average. 

## Part B 

```{r}
X <- as.matrix(cbind(1, X1, X2, X3)) 
Y <- as.matrix(Y)

```
For X'X: 
```{r}
(XX <- t(X) %*% X)
dim(XX)
```
For (X'X)^-1:
```{r}
(XXinv <- solve(XX))
dim(XXinv)
```

For X'Y: 
```{r}
(XY <- t(X) %*% Y)
dim(XY)
```

For b: 

```{r}
(b <- XXinv %*% XY)
dim(b)
```

## Part C

```{r}
MSE <- summary(modela)$sigma^2
(Varb <- MSE * XXinv)

vcov(modela)

```
Var(b0) is `r Varb[1,1]`

Var(b1) is `r Varb[2,2]`

Var(b2) is `r Varb[3,3]`

Var(b3) is `r Varb[4,4]`

Cov(b1,b2) is `r Varb[2, 3]`

Cov(b1,b3) is `r Varb[2, 4]`

Cov(b2,b3) is `r Varb[3, 4]`


vcov and MSE * XXinv gives the same result. 

## Part D

```{r}
anovamodela <- anova(modela)

```
SSR = SumofSq for (age+illness+anxiety)
  = 8275.4 + 480 + 364.2 = 9119.6
SSE = 4248.8
df for regression = p-1. In this case, p =4 so our df is 3
df for error is n-p. In this case, n = 46 so our df is 42. 

For the mean squares, MSR = SSR/(p-1). In this case, MSR = 9119.6/3 = 3039.867. 
For MSE, we calculate it with SSE/(n-p) = 4248.8/42 = 101.1619

Total SS = 13368.4


| Source     | SS      | DF | Mean Squares |
|------------|---------|----|--------------|
| Regression | 9119.6  | 3  | 3039.867     |   
| Error      | 4248.8  | 42 | 101.1619     |   
| Total      | 13368.4 | 45 |              |  


## Part E 

$$H_0 :  \ B_1 =  \ B_2 =  \ B_3 = 0 $$

\[
  H_A: \text{At least one of } \beta_1, \beta_2, \text{ or } \beta_3 \neq 0
\]

$F^*$ = $\frac{MSR}{MSE}$ = $\frac{3039.867}{101.1619}$ = 30.04952

```{r}
Fcrit = qf(0.95,3,42 )
```

F(0.95,3,42) = `r Fcrit`. 

Since $F^*$ > Fcrit  = 30.04952>2.827049, then we reject the null hypothesis and conclude that at least one of the Betas is non-zero.

## Part F

For $b_1$, the $H_0$ is that $b_1$ = 0. 
The $H_A$ is that $b_1 \neq 0$

For $b_2$, the $H_0$ is that $b_2$ = 0. 
The $H_A$ is that $b_2 \neq 0$

For $b_3$, the $H_0$ is that $b_3$ = 0. 
The $H_A$ is that $b_3 \neq 0$

$$
t^* = \frac{b_k}{\sqrt(MSE*(X'X)^{-1})}
$$ 

```{r}
(MSE <- summary(modela)$sigma^2)
# est var of b1
XXinv[2,2]*MSE
b1tval <- b[2]/sqrt(XXinv[2,2]*MSE)
#est var of b2
XXinv[3,3]*MSE
b2tval <- b[3]/sqrt(XXinv[3,3]*MSE)

#est var of b3
XXinv[4,4]*MSE
b3tval <- b[4]/sqrt(XXinv[4,4]*MSE)

#crit t(1-alpha/2,n-p)
critt = qt(0.975,42)

```

For $b_1$, $t^*$ = -1.14/sqrt(0.04613853) = -5.307292
For $b_2$, $t^*$ = -0.44/sqrt(0.2420303) = -0.8943712
For $b_3$, $t^*$ = -13.47/sqrt(50.40518) = -1.897274

The critical t is `r critt`. 

```{r}

b1pval <- 2* pt(b1tval,42)

b2pval <- 2* pt(b2tval,42)
b3pval <- 2* pt(b3tval,42)

```
<!-- Do we need to multiply the pval by 2? -->
Since $|t^*|$ for b1 > critical t, we reject the null hypothesis and conclude that the parameter is nonzero. Similarly, since the pvalue of `r b1pval`<0.05, we can reject the null hypothesis. 

Since $|t^*|$ for b2 and b3 < critical t, we fail to reject the null hypothesis and conclude that there is not enough evidence to determine that the parameter is nonzero.  Similarly, since the pvalue of `r b2pval` and `r b3pval` >0.05, we fail to reject the null hypothesis.


## Part G

```{r}
summary(modela)

```

From the summary function, we can see that the pvalue for the ftest is 1.542e-10 which is less than 0.05, which means we would reject the null hypothesis, like concluded previously, 

The Pvalues for the individual tests match up as well. For b1, we can see the pval is 3.81 * 10^6. This is less than alpha = 0.05 so we reject the null hypothess and conclude that the parameter is nonzero. For the other two, we can see that b2 = 0.37 and b3 = 0.065 which are both greater than 0.05. For these two, the conclusion is that we fail to reject the null hypothesis and that there is not enough evidence to determine that the parameter is nonzero.This is what we concluded. 

## Part H 
$$ r^2 = 1 - \frac{SSE}{SSTO}$$
In this case, the SSE = 4248.8 and the SSTO = 13368.4. Therefore, the r^2 = 0.6821759. This means that 68.22 percent of the variation in the model is explained by the regression function. This value is the same value the summary function gives us. 

## Part I

Individual 95% confidence intervals for B1, B2, B3:

```{r}
alpha = 0.05
n = 46


# 95% Confidence interval for beta1.

lower <- b[2] - qt(p = 1-alpha/2, df = n-4)*sqrt(MSE*XXinv[2,2])
upper <- b[2] + qt(p = 1-alpha/2, df = n-4)*sqrt(MSE*XXinv[2,2])
(b1int <- c(lower, upper))

# 95% Confidence interval for beta2.

lower <- b[3] - qt(p = 1-alpha/2, df = n-4)*sqrt(MSE*XXinv[3,3])
upper <- b[3] + qt(p = 1-alpha/2, df = n-4)*sqrt(MSE*XXinv[3,3])
(b2int <- c(lower, upper))

# 95% Confidence interval for beta3.

lower <- b[4] - qt(p = 1-alpha/2, df = n-4)*sqrt(MSE*XXinv[4,4])
upper <- b[4] + qt(p = 1-alpha/2, df = n-4)*sqrt(MSE*XXinv[4,4])
(b3int <- c(lower, upper))

(confint(modela, level = 0.95))
```

The intervals are the same through manually calculated and using the confint. 

```{r}

alpha_bonf <- alpha / 3
# 95% Bonf Confidence interval for beta1.
lower <- b[2] - qt(p = 1-alpha_bonf/2, df = n-4)*sqrt(MSE*XXinv[2,2])
upper <- b[2] + qt(p = 1-alpha_bonf/2, df = n-4)*sqrt(MSE*XXinv[2,2])
(b1intbon <- c(lower, upper))

# 95% Bonf Confidence interval for beta2.
lower <- b[3] - qt(p = 1-alpha_bonf/2, df = n-4)*sqrt(MSE*XXinv[3,3])
upper <- b[3] + qt(p = 1-alpha_bonf/2, df = n-4)*sqrt(MSE*XXinv[3,3])
(b2intbon <- c(lower, upper))

# 95% Bonf Confidence interval for beta3.
lower <- b[4] - qt(p = 1-alpha_bonf/2, df = n-4)*sqrt(MSE*XXinv[4,4])
upper <- b[4] + qt(p = 1-alpha_bonf/2, df = n-4)*sqrt(MSE*XXinv[4,4])
(b3intbon <- c(lower, upper))

bonlevel <- 1 - 0.05/3

# Bonferroni-adjusted confidence intervals for all coefficients
confint(modela, level = bonlevel)
```
You can see the bonferroni intervals calculated manually and using the confint functions are the same. 

## Part J 

SSTO: 

```{r}
J <- matrix(1, nrow = n, ncol = n)
(SSTO <- t(Y) %*% Y - 1/n*t(Y) %*% J %*% Y)

```

SSR: 
```{r}
(SSR <- t(b) %*% t(X) %*% Y - 1/n*t(Y) %*% J %*% Y)
```

SSE: 

```{r}
(SSE <- t(Y) %*% Y - t(b) %*% t(X) %*% Y)
```


# Question 2 

```{r}
A <- matrix(c(-3, 1, 2, -5), nrow = 2, byrow = TRUE)
B <- matrix(c(0, -2, 1, 4), nrow = 2, byrow = TRUE)
EV <- matrix(c(4, -1), nrow = 2, byrow = TRUE)
EU <- matrix(c(-2, 3), nrow = 2, byrow = TRUE)
VarV <- matrix(c(2, 1,1,3), nrow = 2, byrow = TRUE)
VarU <- matrix(c(1, -1,-1,2), nrow = 2, byrow = TRUE)

```

## Part A

E(AV + BU) = AE[V] + BE[U]

Therefore, E(AV + BU) =

```{r}
(E_AV_BU <- A %*% EV + B %*% EU)
```

## Part B

Var(CX) = CVar(X)C'
Var(C) = 0

So, Var(AV + B) = AVar(X)A' + 0 

So, Var(AV+B) = 

```{r}
(VarAV <- A %*% VarV %*% t(A))
```




## Var C

Var(BU + A) = BVar(U)B' + 0 

The Var(BU + A) is : 
```{r}
(VarBU <- B %*% VarU %*% t(B))

```

